A Few Easy Pieces: \\ How to enhance your relationship with your future self.

### 2011-02-27 workflow.Rnw
Version sent to Wendy and Brian and others for comments.

\documentclass[12pt]{article}
\usepackage{fancyvrb,natbib,url,array,dcolumn,float,booktabs}
\usepackage{hyperref}

@@

\section{Meaningful code requires data.}
All files containing commands operating on data must refer to a data file. A reference to a data file is a line of code the analysis program will use to operate on (``load''/ ``open'' / ``get'' / ``use'') the data file. One should not have to edit this line on different computers or platforms in order to execute this command.  Using R, for example, all analysis files should have load('thedata.rda') or \Verb+read.csv('http://www.mywebsite.org/Data/thedata.csv')+ or some equivalent line in them, and 'thedata.csv' should be stored in some place easy to find (like in the same directory as the file or perhaps in 'Data/thedata.rda').

@@

The principle of modularity \cite{nagler1995coding} suggests that you separate data cleaning, processing, recoding, and merging from analysis in different files. So, perhaps your analysis oriented files will load('cleandata.rda') or something but with a comment in the code telling the future you (among others) that cleandata.rda was created from create-cleandata.R which in turn begins with read.csv('dirtydata.csv') which, from a comment in the file, was downloaded from \url{http://www.greatfreedata.gov/thedata?dirtydata.csv} on April 1, 2011. Such a data processing file will typically end with something like \Verb+save('cleandata.rda')+ so that we are doubly certain about the provenance of the data.

@@

The principle regarding data is to know where the data came from and what operations were performed on which set of data. I have seen command files which do not begin with a call to load data and I have despaired. What is the meaning of \Verb+please-fit(var1 with var2)+ if I do not know where var1 var2 come from.\footnote{The command \Verb+please-fit+ and some other R functions used in this essay come from the MayIPleaseDoStatistics package for R which emphasizes politeness in data analysis. It is highly recommended for use with toddlers.}

\section{Type, don't click.}
All tables and graphs should be generated from code and not clicking (let alone copying and pasting). If I wanted to re-create the figure you created but with red lines, and only for people in the South, I should be able to do so with just a few edits to the code.

@@

<<read.data,echo=FALSE,results=hide>>=

##This read.csv line is slightly more complex than needed because of
##the particular encoding used for characters in the source file.

dem.nations.df<-read.csv(url("http://www.hks.harvard.edu/fs/pnorris/Data/Democracy%20CrossNational%20Data/Democracy%20Crossnational%20Data%20Spring%202009.csv",encoding="ISO_8859-1"),as.is=TRUE)

@@

<<fig1code,echo=TRUE,fig=FALSE,width=3,height=3>>=
##The code to produce the following figure
par(bty="n",xpd=TRUE,pty="s",tcl=-.25)
with(good.df,plot(Gini2004/100,protac2000,
     xlab='Gini Coefficient 2004 (UNDP)',
     ylab='Mean Protest Activities\n(World Values Survey 1980-2000)',
     cex=.8))

with(good.df[c("EGY","JOR","USA","SWE","CHL"),],
     text(Gini2004/100,protac2000,labels=Nation,srt=0,cex=.6,pos=3,offset=.1))
@

\begin{figure}[H]
  \begin{center}
<<fig1plot,echo=FALSE,fig=TRUE,width=3,height=3>>= 
<<fig1code>>
@
    \caption{Protest activity by income inequality \citep[from]{norris2009data}.}\label{fig:giniprot}
  \end{center}
\end{figure}

@@

\section{Enjoy literate data analysis.}
The source code of any pdf document exchanged by the group must be available and executable. Say you run some command and then make a table or a graph. Then you produce a nice little report on your work for use among the working group. Eventually we want to use pieces of that report (tables, graphs, paragraphs) in a publishable paper. Thus, that report must not have been created using copying and pasting. It must be created using LaTeX (or HTML or OpenOffice or something else that the working group standardizes on). And, in fact, it must be created using the reproducible markup system like Sweave (if you are using LaTeX and R as your working group tools) \citep{Leis:2005}. Alternatively, you can put comments into your report detailing where the figures and tables came from (i.e. which command file produced them). Or include details about which code produced which figures in your MANIFEST.txt.

@@

The Sweave system is even nicer because you can have a LaTeX file with R code right inside of it! I showed Sweave with plotting in Figure~\ref{fig:giniplot}, those who view the code for this essay will see how the Table~\ref{tab:protest} was also generated directly from a regression object.

@@

<<printtable,echo=FALSE,results=tex>>= 
print(thextab,include.colnames=FALSE,
      hline.after=NULL,
      table.placement="H",
      add.to.row=list(
        pos=list(0,3),
        command=c(
                  "\\toprule \n &
 \\multicolumn{1}{c}{Coef} &
 \\multicolumn{1}{c}{Std. Err.} &
 \\multicolumn{2}{c}{95\\% CI} \\\\ \\cmidrule(r){2-5} ", 
                  paste("& \\multicolumn{4}{c}{n: ",n,",","   resid. sd:",signif(est.sigma),", R$^2$:",r2,"} \\\\ \\bottomrule",sep="")
                  )
        ))
@

@@

% (the bazaar system appears particularly well suited for use with DropBox http://marcioandreyoliveira.blogspot.com/2009/05/get-your-own-bazaar-hosting-service-for.html)

We use subversion with our own research group, and I use it for all of
my own projects. Subversion and bazaar and git are all great. They
mainly differ in the extent to which you need to run a
server. Subversion requires a server and we are lucky that the
National Center for Supercomputing Applications provides such a server
for us. However, the price of such hosting may not be that much using
one of the many webhosting services out there.

@@

When you reach certain milestones you can rename the file accordingly:
thedocAPSA2009.tex [for the one sent to discussants at APSA] or
thedocAPSR2015.tex [for the version eventually sent to the APSR six
years after you presented it at APSA]. The systems I mentioned above
all allow this kind of thing and are much more elegant and capable,
but you can do it by hand too as long as you don't mind taking up a
lot of disk space and having many many ``thedoc...''  files around.

\section{Write code for your audience.}
Read \citet{nagler1995coding}. Comment your code. Luckily, if you are
using a literate programming practice (Sweave, odfWeave, using R2HTML,
etc..), you can write paragraphs to surround your code as well a
technical comments in the code itself. If you are not using a strictly
literate programming practice then use whatever commenting protocol
exists in your analysis language.

@@

<<reg1,echo=TRUE,results=verb>>=
##  Repeat the regression run for Table~\ref{tab:protest} [use a label instead
##  of a number to make moving sections and tables easier]
##  Do political rights predict protest reporting?
lm1<-lm(protac2000 ~ I(Gini2004/100) + meanpr, good.df)
##  Q: Was it worthwhile controlling for political rights?
lm2<-lm(protac2000~I(Gini2004/100),data=good.df)
coef(lm2)[2]-coef(lm1)[2]
##  A: Yes (around 3/4 an act of protest difference)
@

@@

\section{Unit Testing}

@@

<<unit.test>>=
times.2.fn<-function(thenumber){
  thenumber+2
}

test.times.2.fn<-function(){
  if(times.2.fn(thenumber=4) == 8 &
     times.2.fn(thenumber=-4) == -8){
    print("It works!")
  }else{print("It does not work!")}
}

test.times.2.fn()
@

@@

\section{File Naming}

@@

\section{Exemplars}

Lots of people are thinking about ``reproducible research'' and ``literate programming'' these days. Google those terms. Of course the devil is the details: Here I list a few of my own attempts at enabling reproducible research. You'll find many other inspiring examples on the web.

I have experimented with three systems so far: (1) for one paper we simply included a Sweave document and data files into a compressed archive \cite{bowers2005dataverse}; (2) for another more computing intensive paper we assembled a set of files which enabled reproduction of our results using the ``make'' system \citep{bowers2008dataverse}; and (3) recently I have tried the ``compendium'' approach \citep{gentleman2005reproducible,gentleman2007statistical} which embeds an academic paper with the R package system \cite{bowers2011dataverse}. The benefit of this last system is that one is not required to have access to unix: the compendium is downloadable from within R using \Verb+install.packages()+ and it viewable using the \Verb+vignette()+ function.

\section{Constructive Collaboration}

We all always collaborate. Many of us collaborate with groups of people at one moment in time as we race against a deadline. All of us collaborate with ourselves over time. An idealized version of self-collaboration goes something like this: We begin a project with a few notes. We write a conference paper. We revise this paper and send it to a journal. The journal asks for revisions. We revise and resubmit. Our paper is accepted. We revise a bit more. The paper is published. Later another scholar reacts to our paper. We write another paper based on said reactions (draft, conference paper, submission, revision, acceptance, revision). These processes occur over years of time. So, I collaborate with a self 5 or 10 or even more years hence when I write a paper. [What is a reasonable time-span for which to plan for self-collaboration on a single idea? Ask your advisers how long it took them from idea to dissertation to book (articles).]

The time-frames over which collaboration are required --- whether
among a group of people working together or within a single scholar's
productive life --- are much longer than any given version of any
given software will easily exist. Plain text is the exception. Thus,
even as we extol version control systems, one must have a way to
ensure future access to them in a form that will still be around when
the cockroaches take over political science departments.\footnote{
  (and thereby finally make Ctrl-a Ctrl-x Esc-x Ctrl-c a reasonable
  way to instruct Emacs to do something
  (\url{http://kieran.healy.usesthis.com/} )}

Now, imagine that the UN never get's wind of your work, or even that your article does not spawn debate. Why then would you spend time to communicate with your future self and others? My own answer to this question is because I want my work to be credible and useful.  What I report in my data analyses should have two main characteristics: (1) the findings of the work should not be a matter of opinion and (2) other people could reproduce the findings. That is, what the work represents is not a matter of opinion but is a shared experience --- and an experience shared without respect to the identities of others (although requiring some shared technical training and research resources). Such work should make us change how we act --- or at least, it ought to stand on stronger epistemological ground than other claims about experiences which ought to be shared or shareable.

Assume we want others to believe us when we say something. More narrowly, assume we want other people to believe us when we say something about data: ``data'' here can be words, numbers, musical notes, images, ideas, etc \ldots The point is that we are making some claims about patterns that we see in some collection of stuff. Now, it might be easy to convince other people that ``this collection of stuff is different from this collection of stuff'' if those people were looking over our shoulders the whole time that we made decisions about collecting the stuff and analyzed it [where ``analyze'' means something like ``break it up into understandable parts'']. Unfortunately, we can't assume that people are willing to shadow a researcher throughout her career. Rather, we do our work alone or in small groups and want to convince other distant people about our analyses.

Now, say your collections of stuff are large or complex and where your chosen tools of analyses are computer programs. How can we convince people that what we did with some data with some program is credible: not a matter of whim or opinion, and reproducible by others who didn't shadow us as we wrote our papers? Here are a few concrete actions that people can take to enhance the believeability of their work. In addition, these actions make collaboration within research groups more effective (since believeability comes in part from reproducibility and research groups often need to be able to reproduce in part or in whole what different people in the group have done).

In the end, following these practices allows your analyses of your collections to be credible --- and, in the case of criticism, you can always tell antagonists to go redo your analysis if they really don't believe you.  Let me end with a pithy statement in favor of the idea that collaboration is at the heart of our endeavor:

"[I]f the empirical basis for an article or book cannot be reproduced, of what use to the discipline are its conclusions? What purpose does an article like this serve?'' \cite[445]{king1995replication}

>>>>

\documentclass[12pt]{article}
\usepackage{fancyvrb,natbib,url,array,dcolumn,float,booktabs,listings}
\usepackage{hyperref}

@@

\section{Meaningful code requires data.}

All files containing commands operating on data must refer to a data
file. A reference to a data file is a line of code the analysis
program will use to operate on (``load''/ ``open'' / ``get'' /
                                ``use'') the data file. One should not have to edit this line on
different computers or platforms in order to execute this command.
Using R, for example, all analysis files should have
load('thedata.rda') or
\Verb+read.csv('http://www.mywebsite.org/Data/thedata.csv')+ or some
equivalent line in them, and 'thedata.csv' should be stored in some
place easy to find (like in the same directory as the file or perhaps
                    in 'Data/thedata.rda').

@@

The principle of modularity \cite{nagler1995coding} suggests that you
separate data cleaning, processing, recoding, and merging from
analysis in different files. So, perhaps your analysis oriented files
will load('cleandata.rda') or something but with a comment in the code
telling the future you (among others) that cleandata.rda was created
from create-cleandata.R which in turn begins with
read.csv('dirtydata.csv') which, from a comment in the file, was
downloaded from
\Verb+http://www.greatfreedata.gov/dirtydata.csv+ on April 1,
2011. Such a data processing file will typically end with something
like \Verb+save('cleandata.rda')+ so that we are doubly certain about
the provenance of the data.

@@

The principle regarding data is to know where the data came from and
what operations were performed on which set of data. I have seen
command files which do not begin with a call to load data and I have
despaired. What is the meaning of \Verb+please-fit(var1 with var2)+ if
I do not know where var1 var2 come from.\footnote{The command
                                                  \Verb+please-fit+ and some other R functions used in this essay come
                                                  from the \Verb+MayIPleaseDoStatistics+ package which emphasizes
                                                  politeness in data analysis. The commands \Verb+please-fit+ and
                                                  \Verb+please-plot+ can be blocked and \Verb+would-you-please-fit+ or
                                                  \Verb+may-I-please-have-a-plot+ can be required using
                                                  \Verb+options(politeness=99)+ }

\section{Type, don't click.}

All tables and graphs should be generated from code and not clicking
(let alone copying and pasting). If I wanted to re-create the figure
you created but with red lines, and only for people in the South, I
should be able to do so with just a few edits to the code.

@@

<<read.data,echo=FALSE,results=hide>>=
##  This read.csv line is slightly more complex than needed because of
##  the particular encoding used for characters in the source file.
dem.nations.df<-read.csv(url("http://www.hks.harvard.edu/fs/pnorris/Data/Democracy%20CrossNational%20Data/Democracy%20Crossnational%20Data%20Spring%202009.csv",encoding="ISO_8859-1"),as.is=TRUE)

@@

<<fig1code,echo=TRUE,fig=FALSE>>=
##  Make a scatterplot of Protest by Inequality
with(good.df,plot(Gini2004/100,protac2000,
                  xlab='Gini Coefficient 2004 (UNDP)',
                  ylab='Mean Protest Activities\n(World Values Survey 1980-2000)',
                  cex=.8))

##  Label a few interesting points
with(good.df[c("EGY","JOR","USA","SWE","CHL"),],
     text(Gini2004/100,protac2000,labels=Nation,srt=0,cex=.6,pos=3,offset=.1))
@

\begin{figure}[H]
  \begin{center}
<<fig1plot,echo=FALSE,fig=TRUE,width=3,height=3>>=
par(bty="n",xpd=TRUE,pty="s",tcl=-.25)
<<fig1code>>
@
    \caption{Protest activity by income inequality \citep[from]{norris2009data}.}\label{fig:giniprot}
  \end{center}
\end{figure}

@@

\section{Enjoy literate data analysis.}

The source code of any pdf document exchanged by the group must be
available and executable. Say you run some command and then make a
table or a graph. Then you produce a nice little report on your work
for use among the working group. Eventually we want to use pieces of
that report (tables, graphs, paragraphs) in a publishable paper. Thus,
that report must not have been created using copying and pasting. It
must be created using LaTeX (or HTML or OpenOffice or something else
that the working group standardizes on). And, in fact, it must be
created using the reproducible markup system like Sweave (if you are
using LaTeX and R as your working group tools)
\citep{Leis:2005}. Alternatively, you can put comments into your
report detailing where the figures and tables came from (i.e. which
command file produced them). Or include details about which code
produced which figures in your MANIFEST.txt.

@@

The Sweave system is even nicer because you can have a LaTeX file with
R code right inside of it! I showed Sweave with plotting in
Figure~\ref{fig:giniprot}, those who view the code for this essay will
see how the Table~\ref{tab:protest} was also generated directly from a
regression object.

@@

<<printtable,echo=FALSE,results=tex>>=
  print(thextab,include.colnames=FALSE,
        hline.after=NULL,
        table.placement="H",
        add.to.row=list(
          pos=list(0,3),
          command=c(
            "\\toprule \n &
             \\multicolumn{1}{c}{Coef} &
             \\multicolumn{1}{c}{Std. Err.} &
             \\multicolumn{2}{c}{95\\% CI} \\\\ \\cmidrule(r){2-5} ",
            paste("& \\multicolumn{4}{c}{n: ",n,",",
                  "   resid.sd:",signif(est.sigma),
                  ", R$^2$:",r2,"} \\\\ \\bottomrule",sep="")
            )
          ))
@

@@

We use subversion with our own research group, and I use it for all of
my own projects. Subversion and bazaar and git are all great. They
mainly differ in the extent to which you need to run a
server. Subversion requires a server and we are lucky that the
National Center for Supercomputing Applications provides such a server
for us. However, the price of such hosting may not be that much using
one of the many webhosting services out there. [See Y in this issue
for much more about version control and collaboration.]

@@

 When you reach certain milestones you can rename the file accordingly:
 thedocAPSA2009.tex --- for the one sent to discussants at APSA --- or
 thedocAPSR2015.tex --- for the version eventually sent to the APSR six
 years after you presented it at APSA. The systems I mentioned above
 all allow this kind of thing and are much more elegant and capable,
 but you can do it by hand too as long as you don't mind taking up a
lot of disk space and having many many ``thedoc...''  files around.

\section{No data analyst is an island.}

Read \citet{nagler1995coding}. Comment your code. Luckily, if you are
using a literate programming practice (Sweave, odfWeave, using R2HTML,
etc..), you can write paragraphs to surround your code as well a
technical comments in the code itself. If you are not using a strictly
literate programming practice then use whatever commenting protocol
exists in your analysis language.

@@

<<reg1,echo=TRUE,results=verb>>=

 ##  Repeat the regression run for Table~\ref{tab:protest}
 ##  Do political rights predict protest reporting?
 lm1<-lm(protac2000 ~ I(Gini2004/100) + meanpr, data=good.df)
 ##  Q: Was it worthwhile controlling for political rights?
 lm2<-lm(protac2000~I(Gini2004/100),data=good.df)
 coef(lm2)[2]-coef(lm1)[2]
 ##  A: Yes (around 3/4 an act of protest difference)
@

@@

 Notice one another benefit of coding for an audience: we learn by
 teaching. By assuming that others will look at your code, you will be
 more likely to write clearer code, or perhaps even to learn more about
 what you are doing as you do it.

 \section{Minimize error by testing.}

@@

<<unit.test>>=

  ##  The test function:
  test.times.2.fn<-function(){
    ##  This function tests times.2.fn
    if(times.2.fn(thenumber=4) == 8 &
       times.2.fn(thenumber=-4) == -8){
      print("It works!")
    }else{print("It does not work!")}
  }

##  The function:
times.2.fn<-function(thenumber){
  ##  This function multiplies a scalar number by 2
  ##  thenumber is a scalar number
  thenumber+2
}

##Use the test function
test.times.2.fn()
@

@@

\section{Take advantage of file names to convey information to your future self.}

@@

\section{Copy and improve on others' examples.}

Lots of people are thinking about ``reproducible research'' and
``literate programming'' these days. Google those terms. Of course the
devil is the details: Here I list a few of my own attempts at enabling
reproducible research. You'll find many other inspiring examples on
         the web. Luckily, the open source ethos aligns nicely with academic
         incentives, so we are beginning to find more and more people offering
         their files online for copying and improvement. By the way, if you do
         copy and improve, it is polite to alert the person from whom you made
         the copy about your work.

         I have experimented with three systems so far: (1) for one paper we
         simply included a Sweave document and data files into a compressed
         archive \cite{bowers2005dataverse}; (2) for another more computing
         intensive paper we assembled a set of files which enabled reproduction
         of our results using the ``make'' system \citep{bowers2008dataverse};
         and (3) recently I have tried the ``compendium'' approach
         \citep{gentleman2005reproducible,gentleman2007statistical} which
         embeds an academic paper with the R package system
         \cite{bowers2011dataverse}. The benefit of this last system is that
         one is not required to have access to unix: the compendium is
         downloadable from within R using \Verb+install.packages()+ and it
         viewable using the \Verb+vignette()+ function. The idea that one ought
         to be able to install and run and use an academic paper just as one
         installs and uses statistical software packages is very attractive and
         I anticipate that it will become ever easier to turn papers into R
         packages as creative and energetic folks turn their attention to the
         question of reproducible research.

         \section{Remember that research ought to be credibile communication.}
         \begin{quote}
         [I]f the empirical basis for an article or book cannot
         be reproduced, of what use to the discipline are its conclusions?
         What purpose does an article like this serve?
         \cite[445]{king1995replication}
         \end{quote}

         We all always collaborate. Many of us collaborate with groups of
         people at one moment in time as we race against a deadline. All of us
         collaborate with ourselves over time.\footnote{What is a reasonable
                                                        time-span for which to plan for self-collaboration on a single idea?
                                                        Ask your advisers how long it took them from idea to dissertation to
                                                        book (articles).}
         The time-frames over which collaboration are required --- whether
         among a group of people working together or within a single scholar's
productive life or probably both --- are much longer than any given
version of any given software will easily exist. \href{http://en.wikipedia.org/wiki/Plain_text#Usage}{Plain text} is the
exception. Thus, even as we extol version control systems, one must
have a way to ensure future access to them in a form that will still
be around when sentient cockroaches finally join political science
departments (by then dominated by cetaceans after humans are mostly
uploads).\footnote{The arrival of the six-legged social scientists
  revives Emacs and finally makes Ctrl-a Ctrl-x
  Esc-x Ctrl-c a \href{http://kieran.healy.usesthis.com/}{reasonable
    key combination}.}

Now, imagine that the UN never hears of your work, or even that your
article does not spawn debate. Why then would you spend time to
communicate with your future self and others? My own answer to this
question is because I want my work to be credible and useful to myself
and other scholars even if each article does not immediately change
the world.  What I report in my data analyses should have two main
characteristics: (1) the findings of the work should not be a matter
of opinion and (2) other people could reproduce the findings. That is,
what the work represents is not a matter of opinion but is a shared
experience --- and an experience shared without respect to the
identities of others (although requiring some shared technical
training and research resources).
% Such work should make us change how
% we act --- or at least, it ought to stand on stronger epistemological
% ground than other claims about experiences which ought to be shared or
% shareable.

Assume we want others to believe us when we say something. More
narrowly, assume we want other people to believe us when we say
something about data: ``data'' here can be words, numbers, musical
notes, images, ideas, etc \ldots The point is that we are making some
claims about patterns in some collection of stuff. Now, it
might be easy to convince others that ``this collection of stuff
is different from this collection of stuff'' if those people were
looking over our shoulders the whole time that we made decisions about
collecting the stuff and broke it up into understandable
parts and reorganized and summarized it. Unfortunately, we can't assume that people are willing to
         shadow a researcher throughout her career. Rather, we do our work
         alone or in small groups and want to convince other distant and future people
         about our analyses.

         Now, say your collections of stuff are large or complex and your
         chosen tools of analyses are computer programs. How can we convince
         people that what we did with some data with some program is credible:
           not a matter of whim or opinion, and reproducible by others who didn't
shadow us as we wrote our papers? This essay has suggested a few
concrete ways to enhance the believeability of such scholarly work. In
addition, these actions (as summarized in the section headings of this
essay) make collaboration within research groups more effective.
Believeability comes in part from reproducibility and research groups
often need to be able to reproduce in part or in whole what different
people in the group have done.

In the end, following these practices and those recommended by X and Y
in this issue allows your computerized analyses of your collections
stuff of to be credible.  Finally, if the UN quibbles with your
analyses, your future self can shoot the archive required to reproduce
your work (in still intelligible plain text, analyzed using commented
code so that folks can translate to whatever system succeeds R, or
since you used R, you can include a copy of R and all of the R
packages you used in your final analyses in 2018 in the archive
itself.) You can say, "Here is everything you need to reproduce my
work." To be extra helpful you can add "Read the README file for
futher instructions." And then you can get on with your life: maybe
the next great idea will occur when your 4 year old asks a wacky
question after stripping and painting her overly cooperative 1 year
old brother purple, or teaching a class, or in a coffee shop, or on a
quiet walk.
